---
title: "Day 3 code - Part 2: Data visualization"
subtitle: "BSTA 511/611 Fall 2024 Day 5, OHSU"
author: "Meike Niederhausen, PhD"
date: "10/16/2024"
categories: ["Week 3"]
format: 
  html:
    link-external-newwindow: true
    toc: true
execute:
  echo: true
# editor: visual
editor_options: 
  chunk_output_type: console
---


```{r}
#| label: "setup"
#| include: false
knitr::opts_chunk$set(echo = TRUE, fig.height=3, fig.width=5, message = F)

# run these every time you open Rstudio
library(tidyverse)    
library(oibiostat)
library(ggridges)
library(janitor)
library(rstatix)
library(knitr)
library(gtsummary) # NEW!!
```


# Back to research question

## Case study: discrimination in developmental disability support (1.7.1)

* __Previous research__
  * Researchers examined DDS expenditures for developmentally disabled residents by ethnicity
  * Found that the mean annual expenditures on Hispanics was less than that on White non-Hispanics. 
* __Result__: an allegation of ethnic discrimination was brought against the California DDS.
* __Question: Are the data sufficient evidence of ethnic discrimination?__


## Load `dds.discr` dataset from `oibiostat` package

* The textbook's datasets are in the R package `oibiostat`
* Make sure the `oibiostat` package is installed before running the code below.

* Load the `oibiostat` package and the dataset `dds.discr`

__[the code below needs to be run _every time_ you restart R or render a Qmd file]{style="color:darkorange"}__
```{r}
library(oibiostat)
data("dds.discr")
```

* After loading the dataset `dds.discr` using `data("dds.discr")`, you will see `dds.discr` in the Data list of the Environment window.


### `glimpse()`

__New: `glimpse()`__

* Use `glimpse()` from the `tidyverse` package (technically it's from the `dplyr` package) to get information about variable types. 
* `glimpse()` tends to have nicer output for `tibbles` than `str()`

```{r}
library(tidyverse)
glimpse(dds.discr)  # from tidyverse package (dplyr)
```

## Recall previous data viz

```{r}
#| fig.height: 5
#| fig.width: 8
ggplot(data = dds.discr, 
       aes(x = expenditures,
           y = ethnicity)) + 
  geom_boxplot(color="darkgrey") + 
  labs(x = "Annual Expenditures ($)",
       y = "Race and ethnicity") +
  geom_jitter(     
    aes(color = ethnicity),      
    alpha = 0.3,      
    show.legend = FALSE,      
    position = position_jitter(height = 0.4))      
```

```{r}
#| fig.height: 4
ggplot(data = dds.discr, 
       aes(x = age,
           y = expenditures)) + 
  geom_point() +       
  labs(x = "Age",
       y = "Annual Expenditures ($)") 
```


## Visualize in more detail: ethnicity, age, and expenditures

```{r}
#| fig.height: 8
#| fig.width: 10
dds.discr_Hips_WhnH <- dds.discr %>%  
  filter(ethnicity == "White not Hispanic" | ethnicity == "Hispanic" ) %>% 
  droplevels()   # remove empty factor levels
  

ggplot(data = dds.discr_Hips_WhnH,   
       aes(x = expenditures,
           y = age.cohort)) + 
  geom_boxplot(color="darkgrey") + 
  facet_grid(rows = "ethnicity") +   
  labs(x = "Annual Expenditures ($)",
       y = "Race and ethnicity") +
  geom_jitter(     
    aes(color = ethnicity),      
    alpha = 0.3,      
    show.legend = FALSE,      
    position = position_jitter(     
      height = 0.4))      
```


## Mean annual DDS expenditures by race/ethnicity

### default long format

```{r}
mean_expend <- dds.discr_Hips_WhnH %>% 
  group_by(ethnicity, age.cohort)%>% 
  summarize(ave = mean(expenditures))

mean_expend
```

### wide format

```{r}
mean_expend_wide <- mean_expend %>% 
  pivot_wider(names_from = ethnicity,
              values_from = ave)

mean_expend_wide
```


## Differences in mean annual DDS expenditures by age cohort and race/ethnicity

```{r}
mean_expend_wide <- mean_expend_wide %>% 
  mutate(diff_mean = `White not Hispanic` - Hispanic)

mean_expend_wide
```

__Question__: Are the data sufficient evidence of ethnic discrimination in DDS expenditures when comparing Hispanics with White non-Hispanics?


# Summary of data wrangling so far

* The pipe `%>%` to string together commands in sequence
* `mutate()` to add a new variable to a dataset
* `select()` to select columns (or deselect columns with -variable)
* `filter()` to select specific rows
* `pivot_wider()` to reshape a dataset from a long to a wide format

__Summarizing data__

* `tabyl()` from `janitor` package to make frequency tables of categorical variables
* `summarize()` to get summary statistics of variables
* `group_by()` to group data by categorical variables before finding summaries


